
--- 
title: "README" 
output:  
github_document: 
pandoc_args: ["--wrap=none"] 
always_allow_html: true 
--- 



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Understanding Data Anaylss

  There are many factors that contribute to wine quality. Some of the most important ones to note are weather and climate, balance, and the wine making process.
  We'll start with the first: climate. Before making wine, it is crucial to have high-quality grapes. Depending on the weather, your grapes could either develop an abundance of flavors and juices or end up lacking both. The type of wine someone wants can vary — for example, in cooler climates, grapes tend to be higher in acidity. In hotter climates, grapes tend to ripen more fully, leading to wines with higher sugar content, higher alcohol levels, and a fuller body.
    Balance in wine is also very important. You don't want one element — whether sweetness, acidity, tannin, alcohol, or body — to stand out significantly more than the others. This is why winemakers sometimes let their wine sit for months, or even years, to allow the elements to balance each other out. Lastly, the wine making process itself plays a major role. From the choice of grapes to how those grapes are grown, to fermentation methods and aging periods — every step matters. If one step is not as carefully executed as the others, it can negatively impact the final quality of the wine.


# Data Understanding

Summary 
We have two data sets, one containing data based on red wine and the other based on white wine. There's a combined number of 6,947 observations. 
We have 24 numbers of features between the 2 data sets, 12 for each one.

The features the 2 data sets share are:

Features:
fixed.acidity             pH
free.sulfur.dioxide    quality
volatile.acidity"      citric.acid         
residual.sugar         chlorides            
total.sulfur.dioxide   density                           
sulfates              alcohol             




# Features and  there data types/ descriptions
 
fixed.acidity- Amount of non-volatile acids in the wine (like tartaric acid)|ordinal              
volatile.acidity- Amount of volatile acids that can evaporate and cause sour tastes| ordinal       
citric.acid- Amount of citric acid, which adds freshness and flavor.| ordinal              
residual.sugar- Sugar left over after fermentation (affects sweetness)| ordinal              
chlorides- Salt content in the wine.| ordinal              
free.sulfur.dioxide- Unbound SO₂ that protects wine from spoilage. | ordinal              
total.sulfur.dioxide-Total amount of SO₂ (free + bound) in the wine. | ordinal            
density- How heavy the wine is compared to water. | ordinal              
pH- Measure of wine’s acidity or alkalinity. | Interval           
sulfates- Sulfate salts that help preserve the wine.|ordinal              
alcohol- Percentage of alcohol by volume in the wine | ordinal              
quality- Wine's quality score based on taste tests | Ordinal   

# Data preperations:

I loaded the data, via read.csv, when the data was loaded the  it wasn't in the proper order, the data was separated by ";" rather then "," so to fix that I loaded the data, noting to separated the data with comas rather then ;. 

# Modleing+Evaluating

 Understanding the data's structure*
 
```{r}
#loading data
white_wine_data = read.csv2("winequality-white.csv")
red_wine_data = read.csv2("winequality-red.csv")

#checking rows/col number
nrow(red_wine_data)
nrow(white_wine_data)
ncol(red_wine_data)
ncol(white_wine_data)

```
# Summary Statistics:
```{r}
summary(red_wine_data)
summary(white_wine_data)
```
Red wine summary:

The red wine data set contains 1,599 observations. There are a  few numeric columns ,there's volatile acidity that has a mean of about 0.53, and alcohol ranges from 8.4% to 14.9% with a mean of 10.42%. The wine quality scores range from 3 to 8, with a median of 6 and an average of 5.64.

white wine summary:

The white wine data set has 4,898 observations. Volatile acidity averages 0.28, alcohol ranges from 8.0% to 14.2% with a mean of 10.51%, and pH values average around 3.19. Wine quality scores range from 3 to 9, with a mean of 5.88 and a median of 6.

# checking for missing data
```{r}


colSums(is.na(red_wine_data))
colSums(is.na(white_wine_data))
```
Based on the code, it looks like we have no missing values, in either red or white wine data set. 


# Looking at relationship's within data before creating hypothesis*
 
```{r}
# Histogram of Alcohol in Red and White Wines
red_wine_data$alcohol = as.numeric(red_wine_data$alcohol)
white_wine_data$alcohol = as.numeric(gsub(",", ".", white_wine_data$alcohol))
hist(red_wine_data$alcohol, main = "Alcohol Distribution (Red Wine)", xlab = "Alcohol (%)")
hist(white_wine_data$alcohol, main = "Alcohol Distribution (White Wine)", xlab = "Alcohol (%)")

#Boxplot of alcohol
boxplot(red_wine_data$alcohol, main = "Alcohol Boxplot (Red Wine)", ylab = "Alcohol (%)")
boxplot(white_wine_data$alcohol, main = "Alcohol Boxplot (White Wine)", ylab = "Alcohol (%)")

# Histogram of pH in Red and White Wines
white_wine_data$pH = as.numeric(gsub(",", ".", white_wine_data$pH))
red_wine_data$pH = as.numeric(gsub(",", ".", red_wine_data$pH))
hist(red_wine_data$pH, main = "pH Distribution (Red Wine)", xlab = "pH Level")
hist(white_wine_data$pH, main = "pH Distribution (White Wine)", xlab = "pH Level")

#hist of volatitle
red_wine_data$volatile.acidity = as.numeric(gsub(",", ".", red_wine_data$volatile.acidity))
white_wine_data$volatile.acidity = as.numeric(gsub(",", ".", white_wine_data$volatile.acidity))
hist(red_wine_data$volatile.acidity, main = "Volatile Acidity (Red Wine)", xlab = "Volatile Acidity (g/L)")
hist(white_wine_data$volatile.acidity, main = "Volatile Acidity (White Wine)", xlab = "Volatile Acidity (g/L)")

#Barplot of wine quality
barplot(table(red_wine_data$quality), main = "Quality Scores (Red Wine)", xlab = "Quality Score", ylab = "Count")
barplot(table(white_wine_data$quality), main = "Quality Scores (White Wine)", xlab = "Quality Score", ylab = "Count")
```

# Hypotheses 1
I believe that the higher alcohol content a wine has, the higher the wine quality will be. This is because alcohol can enhance the body and flavor of wine, which are important factors in how quality is perceived." 

# chekcing normality:
```{r}
qqnorm(white_wine_data$alcohol)
qqnorm(red_wine_data$alcohol)
qqline(white_wine_data$alcohol, col = "red")

```

# statistical testing
```{r}
#Test to see if there's correlation between the 2 varibales
cor.test(white_wine_data$alcohol, white_wine_data$quality, method = "spearman")
cor.test(red_wine_data$alcohol, red_wine_data$quality, method = "spearman")
```
# Interpatation:
I first used the Shapiro-Wilk test to check for normality of the alcohol and quality variables for both red and white wines. Since the p-values were significantly less than 0.05, meaning the data was not normally distributed, making parametric methods like Pearson correlation not ideal. So instead I used Spearman correlation instead.
For both wines, the Spearman test returned a p-value < 2.2e-16, and for white wine, a rho of 0.44 — indicating a decent  positive correlation between alcohol content and wine quality. Thus we reject the null hypothesis and find statistically significant support for the idea that higher alcohol levels are associated with higher wine quality.


# Hypothesis 2:
I think that the higher volatile acidity is the lower the  wine's quality is, because Volatile acidity contributes sour, vinegar-like flavors that usually  have a negative effect on the taste and overall perception of wine.

# checking normality:
```{r}
#Checking to see if variable is normally distributed.
qqnorm(white_wine_data$volatile.acidity)
qqnorm(red_wine_data$volatile.acidity)
qqline(red_wine_data$volatile.acidity, col = "blue")

```
# Statistical Testing:
```{r}
#

white_wine_data$volatile.acidity = as.numeric(gsub(",", ".", white_wine_data$volatile.acidity))
white_wine_data$quality = as.numeric(gsub(",", ".", white_wine_data$quality))

#Test to see if there's correlation between the 2 varibales
cor.test(white_wine_data$volatile.acidity, white_wine_data$quality, method = "spearman")
cor.test(red_wine_data$volatile.acidity, red_wine_data$quality, method = "spearman")



```
# Interpertation:
The Spearman test showed a statistically significant negative correlation (rho = -0.197, p < 2.2e-16), which supports the hypothesis in direction. However, the weak strength of the relationship suggests that volatile acidity plays only a small role, so the hypothesis is only partially correct.




# Hypothesis 3:
Due to being fermented longer, which can lead to higher alcohol concentrations compared to white wines. I belive that red wines have a higher average alcohol content than white wines.  

# Checking normality:
```{r}
#Checking to see if variable is normally distributed.
shapiro.test(white_wine_data$alcohol)
shapiro.test(red_wine_data$alcohol)



```
# Statistical testining:
```{r}
 #comparing medians 
wilcox.test(red_wine_data$alcohol, white_wine_data$alcohol)
```
# Interpertaions:
Interpretation for Hypothesis 3:
I first used the Shapiro-Wilk test, which showed that alcohol content in both red and white wines was not normally distributed (p < 2.2e-16 for both). Because the assumptions for a t-test were not met, I used the Wilcoxon rank-sum test, which is a non-parametric alternative. The test returned a p-value of 0.1818, which is greater than 0.05, so we fail to reject the null hypothesis. This suggests there is no statistically significant difference in alcohol content between red and white wines in this data set.

# conclusion:

  In this project,I started by doing a bit of background research to help me get a better understanding with wine and what factors are important to wine quality. By doing so I was able to then perform EDA and statistical tests to see which factors might make wine better or worse in quality.
  First, I thought that wines with higher alcohol content would have higher quality. I checked whether the data was normally distributed, and it wasn’t, so I used a Spearman correlation test instead. The results showed a clear positive relationship between alcohol and wine quality (rho ≈ 0.44, p < 2.2e-16), which supports my idea — higher alcohol levels are linked to better wine quality.
  Next, I tested whether higher volatile acidity would lead to lower wine quality, since it can give wine a sour, vinegar-like taste. The data wasn’t normal here either, so I used another Spearman test. I found a weak but statistically significant negative correlation (rho ≈ -0.20), meaning that more volatile acidity is generally linked to lower wine quality, although the effect isn’t very strong.
  Lastly, I wanted to see if red wine had more alcohol than white wine, based on the idea that red wines are often fermented longer. I used the Wilcoxon rank-sum test because the data wasn’t normal and the groups had different variances. The result (p = 0.1818) showed no significant difference in alcohol content between red and white wines in this dataset, so that hypothesis wasn’t supported.
  In the end, the testing helped show that some things, like alcohol content, really do seem to affect wine quality, while others, like volatile acidity, matter too but not as much. Using EDA and hypothesis testing gave me a better understanding of what betters or worsen wine's quality. 