
--- 
title: "README" 
output:  
github_document: 
pandoc_args: ["--wrap=none"] 
always_allow_html: true 
--- 



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Understanding Data Anaylss

  There are many factors that contribute to wine quality. Some of the most important ones to note are weather and climate, balance, and the wine making process.
  We'll start with the first: climate. Before making wine, it is crucial to have high-quality grapes. Depending on the weather, your grapes could either develop an abundance of flavors and juices or end up lacking both. The type of wine someone wants can vary — for example, in cooler climates, grapes tend to be higher in acidity. In hotter climates, grapes tend to ripen more fully, leading to wines with higher sugar content, higher alcohol levels, and a fuller body.
    Balance in wine is also very important. You don't want one element — whether sweetness, acidity, tannin, alcohol, or body — to stand out significantly more than the others. This is why winemakers sometimes let their wine sit for months, or even years, to allow the elements to balance each other out. Lastly, the winemaking process itself plays a major role. From the choice of grapes to how those grapes are grown, to fermentation methods and aging periods — every step matters. If one step is not as carefully executed as the others, it can negatively impact the final quality of the wine.


# Data Understanding

We have two data sets, one containing data based on red wine and the other based on white wine.   There's a combined number of 6,947 observations. 
We have 24 number of features between the 2 data sets, 12 for each one.

The features the 2 data sets share are:

Features:
fixed.acidity             pH
free.sulfur.dioxide    quality
volatile.acidity"      citric.acid         
residual.sugar         chlorides            
total.sulfur.dioxide   density                           
sulfates              alcohol             

# Features and  there data types/ descriptions
 
fixed.acidity- Amount of non-volatile acids in the wine (like tartaric acid)|Ratio              
volatile.acidity- Amount of volatile acids that can evaporate and cause sour tastes| Ratio       
citric.acid- Amount of citric acid, which adds freshness and flavor.| Ratio              
residual.sugar- Sugar left over after fermentation (affects sweetness)| Ratio              
chlorides- Salt content in the wine.| Ratio              
free.sulfur.dioxide- Unbound SO₂ that protects wine from spoilage. | Ratio              
total.sulfur.dioxide-Total amount of SO₂ (free + bound) in the wine. | Ratio              
density- How heavy the wine is compared to water. | Ratio              
pH- Measure of wine’s acidity or alkalinity. | Interval           
sulfates- Sulfate salts that help preserve the wine.| Ratio              
alcohol- Percentage of alcohol by volume in the wine | Ratio              
quality- Wine's quality score based on taste tests | Ordinal   

# Data preperations:

I loaded the data, via read.csv, when the data was laoded the  it wasnt in the proper order, the data was sperated by ";" rather then "," so to fix that I loaded the data, noting to spperate the data with comas rather then ;. 

# Modleing+Evaluating

 Understanding the data's structure*
 
```{r}
#loading data
white_wine_data = read.csv2("winequality-white.csv")
red_wine_data = read.csv2("winequality-red.csv")

#checking rows/col number
nrow(red_wine_data)
nrow(white_wine_data)
ncol(red_wine_data)
ncol(white_wine_data)

```
# checking for missing data
```{r}


colSums(is.na(red_wine_data))
colSums(is.na(white_wine_data))
```
Based on the code, it looks like we have no missing values, in either red or white wine data set. 


# Looking at relationship's within data before creating hypothesis*
 
```{r}
# Histogram of Alcohol in Red and White Wines
red_wine_data$alcohol = as.numeric(red_wine_data$alcohol)
white_wine_data$alcohol = as.numeric(gsub(",", ".", white_wine_data$alcohol))
hist(red_wine_data$alcohol, main = "Alcohol Distribution (Red Wine)", xlab = "Alcohol (%)")
hist(white_wine_data$alcohol, main = "Alcohol Distribution (White Wine)", xlab = "Alcohol (%)")

#Boxplot of alcohol
boxplot(red_wine_data$alcohol, main = "Alcohol Boxplot (Red Wine)", ylab = "Alcohol (%)")
boxplot(white_wine_data$alcohol, main = "Alcohol Boxplot (White Wine)", ylab = "Alcohol (%)")

# Histogram of pH in Red and White Wines
white_wine_data$pH = as.numeric(gsub(",", ".", white_wine_data$pH))
red_wine_data$pH = as.numeric(gsub(",", ".", red_wine_data$pH))
hist(red_wine_data$pH, main = "pH Distribution (Red Wine)", xlab = "pH Level")
hist(white_wine_data$pH, main = "pH Distribution (White Wine)", xlab = "pH Level")

#hist of volatitle
red_wine_data$volatile.acidity = as.numeric(gsub(",", ".", red_wine_data$volatile.acidity))
white_wine_data$volatile.acidity = as.numeric(gsub(",", ".", white_wine_data$volatile.acidity))
hist(red_wine_data$volatile.acidity, main = "Volatile Acidity (Red Wine)", xlab = "Volatile Acidity (g/L)")
hist(white_wine_data$volatile.acidity, main = "Volatile Acidity (White Wine)", xlab = "Volatile Acidity (g/L)")

#Barplot of wine quality
barplot(table(red_wine_data$quality), main = "Quality Scores (Red Wine)", xlab = "Quality Score", ylab = "Count")
barplot(table(white_wine_data$quality), main = "Quality Scores (White Wine)", xlab = "Quality Score", ylab = "Count")
```

# Hypotheses 1
I believe that the higher alcohol content a wine has, the higher the wine quality will be. This is because alcohol can enhance the body and flavor of wine, which are important factors in how quality is perceived." 

# chekcing assuptions
```{r}
white_wine_data$alcohol = gsub(",", ".", white_wine_data$alcohol)
white_wine_data$alcohol = as.numeric(white_wine_data$alcohol)  

# Checking to see if variable is normally distributed.
shapiro.test(white_wine_data$alcohol)
shapiro.test(red_wine_data$quality)

```
# stitical testing
```{r}

#Test to see if there's correlation between the 2 varibales
cor.test(white_wine_data$alcohol, white_wine_data$quality, method = "spearman")
cor.test(red_wine_data$alcohol, red_wine_data$quality, method = "spearman")
```
# Interpatation:
For both wines, p-value=2.2e-16. Since p-value was signicficantly < then 0.05 for both wines, we reject, thus meaning that there is a statistically significant positive relationship between alcohol content and wine quality.



# Hypothesis 2:
I think that the higher volatile acidity is the lower the  wine's quality is, because Volatile acidity contributes sour, vinegar-like flavors that usually  have a negative effect on the taste and overall perception of wine.

# checking assumptions:
```{r}
#Checking to see if variable is normally distributed.
shapiro.test(white_wine_data$volatile.acidity)
shapiro.test(white_wine_data$quality)

```
# Statistical Testing:
```{r}
#

white_wine_data$volatile.acidity = as.numeric(gsub(",", ".", white_wine_data$volatile.acidity))
white_wine_data$quality = as.numeric(gsub(",", ".", white_wine_data$quality))

#Test to see if there's correlation between the 2 varibales
cor.test(white_wine_data$volatile.acidity, white_wine_data$quality, method = "spearman")
cor.test(red_wine_data$volatile.acidity, red_wine_data$quality, method = "spearman")



```
# Interpertation:
Since the p-values are much smaller than the significance level of 0.05, we reject the null hypothesis in both cases. The negative correlation coefficients suggest that higher volatile acidity is associated with lower wine quality for both red and white wines.  
Thus, the data supports the hypothesis that higher  levels of volatile acidity negativly impacts wine quality.


# Hypothesis 3:
Due to being fermented longer, which can lead to higher alcohol concentrations compared to white wines. I belive that red wines have a higher average alcohol content than white wines.  

# Checking assumptions:
```{r}
#Checking to see if variable is normally distributed.
shapiro.test(white_wine_data$alcohol)
shapiro.test(red_wine_data$alcohol)


#Checking the spread of the two data's
var.test(white_wine_data$alcohol, red_wine_data$alcohol)
```
# Statistical testining*
```{r}
 #comparing medians 
wilcox.test(red_wine_data$alcohol, white_wine_data$alcohol)
```
# Interpertaions*
Interpretation for Hypothesis 3:
At first I used the Shapiro-Wilk test which  indicated that alcohol content in both red and white wines was not normally distributed (p-value < 2.2e-16 for both groups).  
Then the variance test showed that the variances between the two groups were not equal (p-value = 5.947e-12).So I used the non-parametric Wilcoxon rank-sum test to compare the alcohol contents.
- The Wilcoxon rank-sum test produced a p-value of 0.1818.  
Since the p-value is greater than the significance level of 0.05, we fail to reject the null hypothesis. This suggests that there is no statistically significant difference in the average alcohol content between red and white wines in this dataset.

# conclusion

  In this capstone deliverable, I was looking at data sets for white and red wine. I researched a little bit of background information to better help when performing EDA and statistical testing.
  In my testing, I wanted to understand what indicators could be proven to have bettered the quality of wine and indicators that do the opposite (worsen the quality of wine). Based on my understanding of wine and aspects of wine, I hypothesized and tested that the higher alcohol content a wine has, the higher the wine quality will be. This turned out to be true — I know this because the p-value for these statistical tests rejected the null, resulting in there being a correlation between high levels of alcohol and wine quality.
  I also tested if higher volatile acidity lowers wine's quality, because volatile acidity contributes sour, vinegar-like flavors that usually have a negative effect on the taste and overall perception of wine. After testing, I found that, as before, we rejected the null, supporting the idea that there is a strong correlation between high levels of volatile acidity and a decrease in wine quality. There also proved to be a negative correlation coefficient, which suggests that higher volatile acidity is associated with lower wine quality for both red and white wines.
  Lastly, I tested if red wines had higher average alcohol content than white wines, due to being fermented longer, which can lead to higher alcohol concentrations compared to white wines. After testing, it was proved that because the p-value is greater than the significance level of 0.05, we fail to reject the null hypothesis. This suggests that there is no statistically significant difference in the average alcohol content between red and white wines in this dataset.
  Overall, the statistical testing helped confirm that certain chemical aspects like alcohol content and volatile acidity have a measurable impact on wine quality. By using EDA and careful hypothesis testing, I was able to better understand what makes a wine score higher or lower in quality based on these features.



